---
title: "Cyclystics_Data_Analysis"
author: "Ahmed"
date: "08/01/2022"
output:
  pdf_document: default
html_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the libraries and set up directory
```{r libraries & directory}
library(tidyverse)
library(readr)
library(dbplyr)
library(lubridate)
library(rlang)
library(skimr)
library(janitor)
```

dir <- 
  setwd("Dataset/Excel Files")

## Import the last 12 months of csv data files
```{r Import csv files}
td_202101 <- read.csv("202101-divvy-tripdata.csv")
td_202102 <- read.csv("202102-divvy-tripdata.csv")
td_202103 <- read.csv("202103-divvy-tripdata.csv")
td_202104 <- read.csv("202104-divvy-tripdata.csv")
td_202105 <- read.csv("202105-divvy-tripdata.csv")
td_202106 <- read.csv("202106-divvy-tripdata.csv")
td_202107 <- read.csv("202107-divvy-tripdata.csv")
td_202108 <- read.csv("202108-divvy-tripdata.csv")
td_202109 <- read.csv("202109-divvy-tripdata.csv")
td_202110 <- read.csv("202110-divvy-tripdata.csv")
td_202111 <- read.csv("202111-divvy-tripdata.csv")
td_202112 <- read.csv("202112-divvy-tripdata.csv")
```

## Explore each file to make sure column names and types match, if there is a mismatch rbind will throw an error

## Combine the data files into one file
```{r Combine data set into one}
year_cylystic <- rbind(td_202101, td_202102, td_202103, td_202104, td_202105, td_202106, td_202107, td_202108, td_202109, td_202110, td_202111, td_202111,
                       td_202112)
```

## Explore data
```{r Explore size and attributes}
dim(year_cylystic)
head(year_cylystic)
glimpse(year_cylystic)
```
## Check for missing values and delete those rows
```{r NA and delete}
# Change empty spaces to na
year_cylystic[year_cylystic==""]<-NA
# drop the na
year_cylystic <- year_cylystic %>% drop_na()
#check dim for changes
dim(year_cylystic)
summary(year_cylystic)
```

## Comments
The started_at, ended_at columns which represent the start and end times of each ride are shown as character type data, they wil be changed to dttm date type.
Also start_startion_id and end_station_id is shown as character type and will be changed to integer type

```{r Convert data types}
year_cylystic$started_at <- ymd_hms(year_cylystic$started_at)
year_cylystic$ended_at <- ymd_hms(year_cylystic$ended_at)
summary(year_cylystic)
head(year_cylystic)
```

## Comments
We can separate the start_time column into year month day and day can be further broken into day of the week
```{r Break started_at into year, month and day}
year_cylystic <- year_cylystic %>%
  separate(
    started_at, 
    into = c("year", "month", "day"), 
    sep = "-", 
    remove = FALSE
  ) %>%
  separate(
    day, 
    into = c("day_of_week", "time_of_day"), 
    sep = " ",
    remove = FALSE
  )

glimpse(year_cylystic)
```

## Comments
Divide the year into quarters as well as create a day of the week to see if the season and the day of the week affect the rides
```{r Create yearly quarters and create day of the week}
year_cylystic <- year_cylystic %>%
  mutate(
    quarter = cut(
      year_cylystic$started_at, 
      breaks = "quarter", 
      labels = c("Q1", "Q2", "Q3", "Q4"),
      ordered_result = TRUE
    ),
    month = factor(month.name[as.numeric(month)], levels = month.name, ordered = TRUE),
    day_of_week = wday(
      as.Date(year_cylystic$started_at), 
      label = TRUE, 
      abbr = FALSE
    ),
    day = as.numeric(str_sub(day, 1, 2)),
    time_of_day = hms(time_of_day),
  )
```

# Keep distinct ride id's only removing duplicates
```{r Keep distinct ride_id}
year_cylystic %>% 
  distinct(ride_id)
```

```{r create column ride_length}
year_cylystic$ride_length <- difftime(year_cylystic$ended_at, year_cylystic$started_at, units = "mins")
head(year_cylystic)
```

```{r remove bad data}
# data where the rides are less than 1 minute or more than 240 hours (considered stolen)
year_cylystic <- year_cylystic[!(year_cylystic$ride_length < 1 || year_cylystic$ride_length > 14400)]
```

```{r export cleaned data to txt}
write.table(year_cylystic, file = "year_cylystic.txt", sep = "\t",
            row.names = FALSE, col.names = TRUE)
```


```{r Load cleaned data file}
cyclystic <- read.table("year_cylystic.txt",header = TRUE, sep = "\t")
glimpse(cyclystic)
skim(cyclystic)
```

## Analysis

```{r Pie chart of member Vs casual}
pie_user<- cyclystic %>%
  group_by(member_casual) %>%
  summarize(
    total_trips = n(),
  )
pie(pie_user$total_trips,labels = pie_user$member_casual)

ggplot(pie_user, aes(x="", y=total_trips, fill=member_casual)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) + 
  theme_void() # remove background, grid, numeric labels
```

```{r Stats for ride_length}
cyclystic %>%
  summarize(
    mean_trip_duration = mean(ride_length),
    median_trip_duration = median(ride_length),
    min_trip_duration = min(ride_length),
    max_trip_duration = max(ride_length)
  )
```

```{r Stats for ride_length based on clinets member casual}
cyclystic %>%
  group_by(member_casual) %>%
  summarize(
    mean_trip_duration = mean(ride_length),
    median_trip_duration = median(ride_length),
    min_trip_duration = min(ride_length),
    max_trip_duration = max(ride_length),
    total_trips = n(),
  ) %>%
  arrange(-total_trips)
```

# Average ride length for casual members is 32 mins and members 13 mins. Casual ride longer


```{r Stats for ride_length based on day of week and member casual}
ride_type_day <- cyclystic %>%
  group_by(day_of_week, member_casual) %>%
  summarize(
    mean_trip_duration = mean(ride_length),
    median_trip_duration = median(ride_length),
    max_trip_duration = max(ride_length),
    total_trips = n(),
  ) %>%
  arrange(desc(total_trips))
```

```{r Average ride duration per weekday and user type}
ride_type_day %>%
  ggplot(aes(factor(day_of_week,levels= c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                          "Friday", "Saturday", "Sunday")), mean_trip_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Daily number of trips in 2021 per bike type",
    subtitle = "Based on user bike type",
    x = "Day of week",
    y = "Number of trips",
    fill = "User type"
  )
```

```{r Trips per week day per user type graph}
ride_type_day %>%
  ggplot(aes(factor(day_of_week,levels= c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                          "Friday", "Saturday", "Sunday")), total_trips, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Daily number of trips in 2021 per user type",
    subtitle = "Based on user  type",
    x = "Day of week",
    y = "Number of trips",
    fill = "User type"
  )
```

# Higher number of trips for casual on Friday Saturday Sunday, member on Tuesday Wednesday


## Graph based on tips per hour of day

# analyzing ridership data by ride type and month

```{r Ride typer per month}
ride_type_month <- cyclystic %>%
  group_by(rideable_type, member_casual,month) %>%
  summarize(
    number_of_trips = n(),
    average_duration = mean(ride_length), 
    median_duration = median(ride_length), 
    max_trip_duration = max(ride_length)
  ) %>%
  arrange(member_casual,rideable_type)

ride_type_month
```

```{r Ride typer per month graph}
ride_type_month %>%
  ggplot(aes(factor(month,levels= c("January","February","March","April","May","June",
                                    "July","August","September","October","November","December")),number_of_trips, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "The monthly number of trips based on user type per month",
    subtitle = "Based on user type: casual and member",
    x = "Month",
    y = "Number of trips",
    fill = "User type"
  ) 
```

# Casual members peak in the summer months June to September

# analyzing ridership data by ride type and weekday

```{r Ride type per week day}
ride_type <- cyclystic %>%
  group_by(rideable_type, member_casual,day_of_week) %>%
  summarize(
    number_of_trips = n(),
    average_duration = mean(ride_length), 
    median_duration = median(ride_length), 
    max_trip_duration = max(ride_length)
  ) %>%
  arrange(member_casual,rideable_type)

ride_type
```

# visualizing the daily number of trips by user type

```{r Ride typer per week day graph}
ride_type %>%
  ggplot(aes(factor(day_of_week,levels= c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                          "Friday", "Saturday", "Sunday")), number_of_trips, fill = rideable_type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Daily number of trips in 2021 per bike type",
    subtitle = "Based on user bike type",
    x = "Day of week",
    y = "Number of trips",
    fill = "User type"
  )
```

# Classic bikes much more popular across al days followed by electric bikes


# analyzing ridership data by user type and weekday

```{r User type per weekday}
day_type <- cyclystic %>%
  group_by(day_of_week, member_casual) %>%
  summarize(
    number_of_trips = n(),
    average_duration = mean(ride_length), 
    median_duration = median(ride_length), 
    max_trip_duration = max(ride_length)
  ) %>%
  arrange(member_casual,day_of_week)

day_type
```


# visualizing the daily number of trips by user type

```{r User type per weekday graph}
day_type %>%
  ggplot(aes(factor(day_of_week,levels= c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                          "Friday", "Saturday", "Sunday")), number_of_trips, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Daily number of trips in 2021",
    subtitle = "Based on user type: member and casual",
    x = "Day of week",
    y = "Number of trips",
    fill = "User type"
  )
```


# visualizing daily average of trip duration by user type

```{r Average duration pwer weekday and user type}
day_type %>%
  ggplot(aes(factor(day_of_week,levels= c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                          "Friday", "Saturday", "Sunday")), average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Daily average of trip duration in 2021",
    subtitle = "Based on user type: member and casual",
    x = "Day of week",
    y = "Average duration (min)",
    fill = "User type"
  )
```


# visualizing the daily number of trips by user type

```{r Average duration per weekday and user type}
day_type %>%
  ggplot(aes(factor(day_of_week,levels= c("Monday", "Tuesday", "Wednesday", "Thursday", 
                                          "Friday", "Saturday", "Sunday")), median_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Daily median of trip duration in 2021",
    subtitle = "Based on user type: customer and subscriber",
    x = "Day of week",
    y = "Median of trip duration (min)",
    fill = "User type"
  )
```


```{r Monthly stats for rides }
month_type <- cyclystic %>%
  group_by(month, member_casual) %>%
  summarize(
    number_of_trips = n(),
    average_duration = mean(ride_length), 
    median_duration = median(ride_length), 
    max_trip_duration = max(ride_length)
  ) %>%
  arrange(member_casual,month)

month_type
```

# visualizing the daily number of trips by user type

```{r Monthly user type and number of trips}

month_type %>%
  ggplot(aes(factor(month,levels= c("January","February","March","April","May","June",
                                    "July","August","September","October","November","December")),number_of_trips, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "The monthly number of trips based on user type",
    subtitle = "Based on user type: casual and member",
    x = "Month",
    y = "Number of trips",
    fill = "User type"
  ) 

```

```{r Monthly user type and number of trips graph}
month_type %>%
  ggplot(aes(factor(month,levels= c("January","February","March","April","May","June",
                                    "July","August","September","October","November","December")),median_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "The monthly median duration based on user type",
    subtitle = "Based on user type: casual and member",
    x = "Month",
    y = "Number of trips",
    fill = "User type"
  ) 
```


```{r Average ride duration per month graph}

month_type %>%
  ggplot(aes(factor(month,levels= c("January","February","March","April","May","June",
                                    "July","August","September","October","November","December")),average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "The monthly average duration based on user type",
    subtitle = "Based on user type: casual and member",
    x = "Month",
    y = "Number of trips",
    fill = "User type"
  ) 
```


```{r Member type per quarter}
quarter_type <- cyclystic %>%
  group_by(quarter, member_casual) %>%
  summarize(
    number_of_trips = n(),
    average_duration = mean(ride_length), 
    median_duration = median(ride_length), 
    max_trip_duration = max(ride_length)
  ) %>%
  arrange(member_casual,quarter)

quarter_type
```

# number of trips per quarter
```{r Number of trips per quarter}
ggplot(quarter_type, aes(quarter, number_of_trips, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Quarterly total trips based on user type",
    subtitle = "Based on user type: casual and member",
    x = "Quarter",
    y = "Total trips",
    fill = "User type"
  )
```


# Average ride duration per quarter
```{r Average ride duration per quarter graph}
ggplot(quarter_type, aes(quarter, average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Quarterly average duration based on user type",
    subtitle = "Based on user type: casual and member",
    x = "Quarter",
    y = "Total trips",
    fill = "User type"
  )
```


# number of trips per quarter
```{r Median duration per quarter graph}
ggplot(quarter_type, aes(quarter, median_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Quarterly median duration based on user type",
    subtitle = "Based on user type: casual and member",
    x = "Quarter",
    y = "Total trips",
    fill = "User type"
  )

```

